Impact of grids and dynamical cores in CESM2.2 on the surface mass balance of the Greenland Ice Sheet

### 2022-05-05 arcticgrids.tex
paper updates

@@

The accuracy of the simulated SMB is expected to be sensitive to grid resolution. Figure~\ref{fig:grisdx}b shows the average grid spacing over GrIS in all six grids in this study. The $\tt ne30pg2$ grid has the coarsest representation with an average grid spacing ($\Delta x$) of $\Delta x=160~km$, and the $\tt Arctic-GrIS$ grid has the highest resolution with an average grid spacing of $\Delta x=14.6~km$, broadly similar to the $11~km$ grid spacing of the RACMO2.3 grid. The $\tt ne30pg3$ grid has an average $\Delta x=111.2~km$, substantially coarser than the $\tt f09$ grid, with an average $\Delta x=60~km$. Although $\tt ne30pg3$ and $\tt f09$ have similar average grid spacing over the entire globe, and comparable computational costs, the convergence of meridians on the FV grid enhances the resolution over the GrIS. The $\tt Arctic$ grid has an average grid spacing of $\Delta x=27.8~km$, and is about 10 times more expensive than the $1^{\circ}$ models.  The $\tt Arctic-GrIS$ grid is about twice as expensive as the $\tt Arctic$ grid.

@@

Some of these temperature differences may be related to different summer shortwave cloud forcing for the various grids and dycores. Figure~\ref{fig:SWCF} shows the summer shortwave cloud forcing bias in six runs, using the CERES-EBAF product. A negative bias corresponds to excessive reflection and cooling. The uniform grids have similar biases, with the clouds reflecting 20--40 W/m$^2$ too much shortwave radiation over a wide swath of the Arctic, primarily the land masses. There is also a halo of positive bias (clouds not reflective enough) around the ocean perimeter of Greenland. The $\tt Arctic$ grid has much smaller cloud forcing biases over the Arctic land masses, but is still too reflective over Alaska, the Canadian Archipelago, and parts of Eurasia.  Compared to the $\tt Arctic$ grid, the $\tt Arctic-GrIS$ grid vastly reduces the cloud forcing bias over Eurasia, and also improves the bias over North America. In both VR grids, the halo of positive shortwave cloud forcing bias around the perimeter of Greenland is absent.

@@

\subsection{Shortwave radiation over Greenland}

In addition to summer temperatures, shortwave radiation is an important determinant of snow and ice melt. Figure~\ref{fig:FSDS} shows the summer incident shortwave radiation bias at the surface over Greenland and surrounding seas. The top panel shows the bias relative to the CERES-EBAF dataset, and the bottom panel relative to the RACMO2.3p2 dataset.
The halo of excessive incident shortwave radiation around the coasts of Greenland is apparent for both datasets in relation to the coarser grids, consistent with the shortwave cloud forcing biases in Figure~\ref{fig:SWCF}.

The ice sheet interior receives too little shortwave radiation on the coarser grids. On the VR grids, both the interior shortwave deficit and the excessive shortwave around the ocean perimeter are improved. This suggests that the coarse-grid clouds are too thick in the Greenland interior and too thin around the perimeter, and that increasing horizontal resolution reduces these biases. This is consistent with the total summer cloud fraction bias, computed from the CALIPSO-GOCCP cloud dataset  and shown in Figure~\ref{fig:prect}. Note that total cloud fraction characterizes the cloud field at all vertical levels, and attenuates the changes arising from any single layer due to the occurrence of overlapping clouds at all other levels. Despite the attenuated signal, the total cloud fraction for the VR grids does indicate reduced cloud coverage in the interior and increased cloudiness around the ocean perimeter.

The agreement of the cloud biases in and around Greenland from multiple independent datasets shows that the biases are a robust feature of the coarser grids. The reduced biases on the VR grids suggest that the coarse-grid biases are a result of insufficient horizontal resolution.

\begin{figure}[t]
\begin{center}
         \includegraphics[width=130mm]{figs/temp_contours_diffCERESdiffRACMO_FSDS.pdf}
\end{center}
\caption{1979-1998 northern hemisphere summer, incident shortwave radiation bias, computed as the difference (top) from CERES-EBAF, and (bottom) RACMO2.3p2 dataset. The differences in the top panel are found by mapping the model output to the $1^{\circ}$ CERES-EBAF grid, and differences on the bottom panel are computed after mapping the RACMO2.3p2 dataset to the individual model grids. Note that the averaging period for the CERES-EBAF panels, 2003-2020, is different from the averaging period for the model results.}
\label{fig:FSDS}
\end{figure}

\begin{figure}[t]
\begin{center}
         \includegraphics[width=130mm]{figs/temp_contours_diffCERESdiffRACMO_CLOUD_PRECIP.pdf}
\end{center}
\caption{1979-1998 northern hemisphere summer (top) total cloud fraction bias, relative to the CALIPSO-GOCCP dataset, and (bottom) precipitation rate bias, relative to the RACMO2.3p2 dataset. The CALIPSO-GOCCP differences are computed after mapping all model output to the $1^{\circ}$ grid, whereas the RACMO differences are computed after mapping the RACMO dataset to the individual model grids. Note that the averaging period for the CALIPSO-GOCCP panels (2006-2017) is different than the model averaging period. \color{red}{ARH - still trying to fix the layout of this figure so the label bars aren't on top of the panels}.}
\label{fig:prect}
\end{figure}

@@

The lower panels of Figure~\ref{fig:prect} show the summer climatological mean precipitation bias over the GrIS, expressed as the fractional difference from the RACMO2.3p2 solution. The coarse $1^{\circ}-2^{\circ}$ grids have large, positive biases centered over the southern dome. The $\tt Arctic$ grid reduces this bias substantially, and the $\tt Arctic-GrIS$ grid reduces it further. This suggests that the southern dome bias arises from inadequate horizontal resolution, and consistent with the positive GrIS accumulation biases that are common to low resolution GCMs \citeA{P2000GPC,VETAL2018TC}.

Large GrIS accumulation rates result from synoptic systems arriving primarily from the south. These systems are orographically lifted at the ice sheet margin, especially over the steep slopes in southeast Greenland (dominated by winter precipitation), concentrating heavy precipitation near the ice margin. At lower resolutions, the topography is too smooth to generate a realistic orographic uplift, leading to anomalously high SMB in the GrIS interior. On the higher resolution grids, precipitation centers migrate from the interior toward the margins, and even out over the ocean in the case of the $\tt Arctic-GrIS$ grid. The more accurate representation of orographic precipitation on the higher resolution grids is consistent with the discussion on simulated cloud and radiation biases, cf. Figures~\ref{fig:SWCF}, \ref{fig:FSDS}, and \ref{fig:prect}.

@@

To illustrate the regional behavior of the SMB components, Figure~\ref{fig:basin} shows the precipitation and combined snow/ice melt integrated over the basins defined by \citeA{RM2012GRL}. The uncertainty due to differences in basin area is larger than for GrIS-wide integrals, owing to the differences in basin boundaries as represented by the common ice masks, which are shown in the $\tt f19$ and $\tt ne30pg2$ panels of Figure~\ref{fig:melt}. Nonetheless, the regional totals in Figure~\ref{fig:basin} correctly show the southeast and southwest basins have the most accumulation. In all basins, accumulation decreases monotonically with increasing grid resolution, though with some exceptions. The $\tt Arctic-GrIS$ grid simulates less precipitation than RACMO in the central-east and southeast basins, and is closest of all grids to the RACMO precipitation in the large southwest basin.

\begin{figure}[t]
\begin{center}
         \includegraphics[width=100mm]{figs/temp_tseries_BASIN.pdf}
\end{center}
\caption{1979--1998 basin integrated components of the SMB; (top) precipitation, (middle) ice/snow melt and (bottom) ice/snow melt estimated from the PDD method. Whiskers span the max/min of the four ensemble members generated from the common-ice-mask approach. Basin definitions are after \citeA{RM2012GRL}, and are found on the common ice masks using a nearest neighbor approach, and shown in Figure~\ref{fig:melt}.}
\label{fig:basin}
\end{figure}

\begin{figure}[t]
\begin{center}
         \includegraphics[width=130mm]{figs/temp_contours_diffRACMO_melt.pdf}
\end{center}
\caption{1979--1998 ice/snow melt biases (in mm/yr) relative to RACMO2.3p2, evaluated on the native model grids. The \citeA{RM2012GRL} basin boundaries are shown in grey for each model grid. Note that Figure~\ref{fig:basin} uses the basin boundaries for the two common ice masks, shown in the $\tt f19$ and $\tt ne30pg2$ panels, in computing the basin-scale integrals. Blue lines in the $\tt f19$ panel show the location of the two transects plotted in Figure~\ref{fig:ztrans}.}
\label{fig:melt}
\end{figure}

@@

Figure~\ref{fig:melt} presents the biases in the combined ice/snow melt as map plots. These plots show that the largest melt biases are on the southeast and northwest coasts, where large coarse-grid cells overlap with the ocean. One possibility is that these problematic grid cells are situated at lower elevations than the true ice sheet surface, leading to a warm bias and too much melt. Figure~\ref{fig:ztrans} shows the representation of the ice sheet surface along two transects on the different grids, compared to the high-resolution dataset used to generate CAM topographic boundary conditions \cite{GMTED2010,gmdd-8-4623-2015}. The two transects are shown in Figure~\ref{fig:melt}: the east-west ``K-transect'' in southwest Greenland and a transect extending from the central dome down to the  Kangerlussuaq glacier on the southeast coast. The $1^{\circ}-2^{\circ}$ grids are noticeably coarse, with only a few grid cells populating the transect. The $\tt f09$ grid is a bit of an exception for the K-transect, with grid cells becoming narrow in the meridional direction at high latitudes. The VR grids are more skillful at reproducing the steep margins of the ice sheet, capturing the parabolic shape of the GrIS margins.

>>>>

The accuracy of the simulated surface mass balance is expected to be sensitive to grid resolution. Figure~\ref{fig:grisdx}b shows the average grid spacing over the Greenland Ice Sheet (\textit{GrIS} hereafter) in all six grids in this study. The $\tt ne30pg2$ grid has the coarsest representation with an average grid spacing ($\Delta x$) of $\Delta x=160$ km, and the $\tt Arctic-GrIS$ grid has the highest resolution with an average grid spacing of $\Delta x=14.6$ km, similar to the $11$ km grid spacing of the RACMO2.3 grid. The $\tt ne30pg3$ grid has an average $\Delta x=111.2$ km, substantially coarser than the $\tt f09$ grid, with an average $\Delta x=60$ km. Although $\tt ne30pg3$ and $\tt f09$ have similar average grid spacing over the entire globe, and comparable computational costs, the convergence of meridians on the FV grid enhances the resolution over the GrIS. The $\tt Arctic$ grid has an average grid spacing of $\Delta x=27.8$ km, and is about 10 times more expensive than the $1^{\circ}$ models.  The $\tt Arctic-GrIS$ grid is about twice as expensive as the $\tt Arctic$ grid.

@@

@@

Some of these temperature differences may be related to different summer shortwave cloud forcing for the various grids and dycores. Figure~\ref{fig:SWCF} shows the summer shortwave cloud forcing bias in six runs, using the CERES product. A negative bias corresponds to excessive reflection and cooling. The uniform grids have similar biases, with the clouds reflecting 20--40 W/m$^2$ too much shortwave radiation over a wide swath of the Arctic, primarily the land masses. There is also a halo of positive bias (clouds not reflective enough) around the ocean perimeter of Greenland. The $\tt Arctic$ grid has much smaller cloud forcing biases over the Arctic land masses, but is still too reflective over Alaska, the Canadian Archipelago, and parts of Eurasia.  Compared to the $\tt Arctic$ grid, the $\tt Arctic-GrIS$ grid vastly reduces the cloud forcing bias over Eurasia, and also improves the bias over North America. In both VR grids, the halo of positive shortwave cloud forcing bias around the perimeter of Greenland is absent.

@@

\subsection{Clouds and precipitation over Greenland}

In addition to summer temperatures, shortwave radiation is an important determinant of snow and ice melt. Figure~\ref{fig:FSDS} shows the summer incident shortwave radiation bias at the surface over Greenland and surrounding seas. The middle panel shows the bias relative to the CERES dataset, and the bottom panel relative to the RACMO2.3p2 dataset. The halo of excessive incident shortwave radiation around the coasts of Greenland is apparent for both datasets in relation to the coarser grids, consistent with the shortwave cloud forcing biases in Figure~\ref{fig:SWCF}.

The ice sheet interior receives too little shortwave radiation in the coarser grids. On the VR grids, both the interior shortwave deficit and the excessive shortwave around the ocean perimeter are improved. This suggests that the coarse-grid clouds are too thick in the Greenland interior and too thin around the perimeter. This is consistent with the total summer cloud fraction bias, computed from the CALIPSO cloud dataset and shown in the top panel of Figure~\ref{fig:FSDS}. Note that total cloud fraction characterizes the cloud field at all vertical levels, and attenuates the changes arising from any single layer due to the occurrence of overlapping clouds at all other levels. Despite the attenuated signal, the cloud fractions are too high in the interior and and too low around the oceanic perimeter, in the coarser grids, while the VR grids have an overall reduction in total cloud fraction bias.

The top panel of Figure~\ref{fig:prect} show the annual climatological mean precipitation bias over the GrIS, expressed as the fractional difference from the RACMO2.3p2 solution. The coarse $1^{\circ}-2^{\circ}$ grids have large, positive biases centered over the southern dome. The $\tt Arctic$ grid reduces this bias substantially, and the $\tt Arctic-GrIS$ grid reduces it further, with precipitation centers migrating from the interior toward the margins.

The more accurate representation of orographic precipitation on the higher resolution grids is consistent with the cloud and radiation biases, cf. Figures~\ref{fig:SWCF}, \ref{fig:FSDS}, and \ref{fig:prect}. The agreement of the cloud and precipitation biases in and around Greenland from multiple independent datasets indicates that the biases are a robust feature of the coarser grids. The reduced biases on the VR grids suggest that the biases in the coarse models are a result of insufficient horizontal resolution.

\begin{figure}[t]
\begin{center}
         \includegraphics[width=130mm]{figs/temp_contours_diffCALIPSOdiffCERESdiffRACMO.pdf}
\end{center}
\caption{1979-1998 northern hemisphere summer (top) total cloud fraction bias relative to the CALIPSO dataset and incident shortwave radiation bias (W/m2), computed as the difference (middle) from CERES, and (bottom) RACMO2.3p2 dataset. The CALIPSO and CERES differences are found by mapping the model output to the $1^{\circ}$ grid, and differences in the bottom panel are computed after mapping the RACMO2.3p2 dataset to the individual model grids. Note that the averaging period for the CALIPSO-GOCCP and CERES-EBAF panels, 2006-2017 and 2003-2020, respectively, are different from the averaging period for the model results.}
\label{fig:FSDS}
\end{figure}

\begin{figure}[t]
\begin{center}
         \includegraphics[width=130mm]{figs/temp_contours_diffRACMO}
\end{center}
\caption{1979-1998 (top) annual precipitation and (bottom) ice/snow melt biases relative to RACMO2.3p2, evaluated on the native model grids. The precipitation biases are expressed as fractional changes, whereas the melt biases are absolute changes (mm/yr). In the bottom panel, the \citeA{RM2012GRL} basin boundaries are shown in grey for each model grid. Note that Figure~\ref{fig:basin} uses the basin boundaries for the two common ice masks, shown in the $\tt f19$ and $\tt ne30pg2$ panels, in computing the basin-scale integrals. Blue lines in the $\tt f19$ panel show the location of the two transects plotted in Figure~\ref{fig:ztrans}..}
\label{fig:prect}
\end{figure}

@@

@@

To illustrate the regional behavior of the SMB components, Figure~\ref{fig:basin} shows the precipitation and combined snow/ice melt integrated over the basins defined by \citeA{RM2012GRL}. The uncertainty due to differences in basin area is larger than for GrIS-wide integrals, owing to the differences in basin boundaries as represented by the common ice masks, which are shown in the $\tt f19$ and $\tt ne30pg2$ panels of Figure~\ref{fig:prect}. Nonetheless, the regional totals in Figure~\ref{fig:basin} correctly show the southeast and southwest basins have the most accumulation. In all basins, accumulation decreases monotonically with increasing grid resolution, though with some exceptions. The $\tt Arctic-GrIS$ grid simulates less precipitation than RACMO in the central-east and southeast basins, and is closest of all grids to the RACMO precipitation in the large southwest basin.

\begin{figure}[t]
\begin{center}
         \includegraphics[width=100mm]{figs/temp_tseries_BASIN.pdf}
\end{center}
\caption{1979--1998 basin integrated components of the SMB; (top) precipitation, (middle) ice/snow melt and (bottom) ice/snow melt estimated from the PDD method. Whiskers span the max/min of the four ensemble members generated from the common-ice-mask approach. Basin definitions are after \citeA{RM2012GRL}, and are found on the common ice masks using a nearest neighbor approach, and shown in Figure~\ref{fig:prect}.}
\label{fig:basin}
\end{figure}

@@

Figure~\ref{fig:prect} presents the biases in the combined ice/snow melt as map plots. These plots show that the largest melt biases are on the southeast and northwest coasts, where large coarse-grid cells overlap with the ocean. One possibility is that these problematic grid cells are situated at lower elevations than the true ice sheet surface, leading to a warm bias and too much melt. Figure~\ref{fig:ztrans} shows the representation of the ice sheet surface along two transects on the different grids, compared to the high-resolution dataset used to generate CAM topographic boundary conditions \cite{GMTED2010,gmdd-8-4623-2015}. The two transects are shown in Figure~\ref{fig:prect}: the east-west ``K-transect'' in southwest Greenland and a transect extending from the central dome down to the  Kangerlussuaq glacier on the southeast coast. The $1^{\circ}-2^{\circ}$ grids are noticeably coarse, with only a few grid cells populating the transect. The $\tt f09$ grid is a bit of an exception for the K-transect, with grid cells becoming narrow in the meridional direction at high latitudes. The VR grids are more skillful at reproducing the steep margins of the ice sheet, capturing the parabolic shape of the GrIS margins.
